### 2017-04-02 paper/oopsla17-submission/learning.tex
some intro
Wes pass over Section 3

\section{Learning to Blame}
\label{sec:learning}
Next, we describe our approach to localizing type errors, in the context
of \lang (\autoref{fig:syntax}), a simple lambda calculus with integers,
booleans, pairs, and lists.
%
\input{syntax}
%
First, we define the inputs to the model, a set of \emph{features}
(\autoref{sec:features}) that we will use to describe programs.
%
Second, we define the expected outputs of the model, a set of
\emph{labels} (\autoref{sec:labels}) that we will use to assign blame.
%
Third, we describe the actual \emph{models} (\autoref{sec:models}) that we
will evaluate.

@@

\begin{figure}[ht]
\begin{minipage}{0.6\linewidth}
\begin{code}
  let rec sumList xs =
    match xs with
    | []     -> []
    | hd::tl -> hd + sumList tl
\end{code}
\end{minipage}
\begin{minipage}{0.3\linewidth}
\begin{code}[numbers=left]
[]
hd + sumList tl
sumList tl
tl
\end{code}
\end{minipage}
\caption{An ill-typed program (left) and a selection of its
  sub-expressions (right).}
\label{fig:sumList}
\end{figure}
\begin{table}[ht]
\begin{tabular}{lrrrrrr}
\toprule
Expression
  & \IsNil & \IsCaseListP & \ExprSize
  & \HasTypeIntCOne & \HasTypeList & \InSlice \\
\midrule
|[]|
  & 1 & 1 & 1 & 0 & 1 & 1 \\
|hd + sumList tl|
  & 0 & 1 & 5 & 1 & 0 & 1 \\
|sumList tl|
  & 0 & 0 & 3 & 0 & 1 & 1 \\
|tl|
  & 0 & 0 & 1 & 0 & 1 & 0 \\
\bottomrule
\end{tabular}
\caption{A selection of the features we would extract from the
  \lstinline!sumList! program in \autoref{fig:sumList}. A feature is
  considered \emph{enabled} if it has a non-zero value, and
  \emph{disabled} otherwise. A ``-P'' suffix indicates that the feature
  describes the parent of the current expression, a ``-C$n$'' suffix
  indicates that the feature describes the $n$-th (left-to-right) child
  of the current expression.  Note that, since we rely on a partial
  typing derivation, we are subject to the well-known traversal bias and
  label the expression \lstinline!sumList tl! as having type
  $\tlist{\cdot}$. The model will have to learn to correct for this
  bias.}
\label{tab:sumList}
\end{table}

@@

We distinguish between matching on a list vs.\ on a pair, as this
affects the typing derivation.
%
We also assume that all pattern matches are well-formed --- \ie all
patterns must match on the same type.
%
Ill-formed match expressions would lead to a type error; however, they
are already effectively localized to the match expression itself.
%
This is not a \emph{fundamental} limitation though, one could easily add
features that specify whether a match \emph{contains} a particular
pattern, and thus have a match expression that enables multiple
features.

\subsubsection{Contextual syntactic features}
These are like local syntactic features, but lifted to describe the
parent and children of the current expression.
%
For example, the \IsCaseListP feature in \autoref{tab:sumList} describes
whether current expression's \emph{parent} matches on a list.
%
If a particular $e$ does not have children (\eg a variable $x$) or a
parent (\ie the root expression), we leave the corresponding features
disabled.
%
This gives us a notion of the \emph{context} in which an expression
occurs, similar to the \emph{n-grams} commonly found in linguistic
models.

@@

\subsubsection{Expression size}
We also add a feature representing the \emph{size} of each expression,
\ie how many sub-expressions does it contain?
%
For example, the \ExprSize feature in \autoref{tab:sumList} is set to 3
for the expression |sumList tl| as it contains 3 expressions:
the two variables and the application itself.
%
This allows the model to learn that, \eg, expressions closer to the
leaves are more likely to be blamed than expressions closer to the root.

@@

Note that our use of typing features in an \emph{ill-typed} program
subjects us to the infamous traversal
bias~\cite{McAdam1998-ub,Yang2000-yr} --- in particular with the
|sumList tl| expression, which might alternatively be assigned
the type $\tint$.
%
We rely on the model to counteract this bias.

\subsubsection{Type error slice}
Finally, we would like the model to be able to distinguish between
changes that could fix the error, and changes that
\emph{cannot possibly} fix the error.
%
Thus, we compute a minimal type error \emph{slice} for the program
(\ie the set of expressions that contribute to the error), and add a
feature that is enabled \emph{iff} an expression is part of the slice.
%
The \InSlice feature in \autoref{tab:sumList} indicates whether an
expression is part of such a minimal slice, and is enabled for all of
the sampled expressions except for |tl|, which does not affect
the type error.
%
If the program contains multiple type errors, we will compute
a minimal slice for each error.

@@

\subsection{Learning Algorithms}
\label{sec:models}
\ES{TODO: mention that this is a SUPERVISED, CLASSIFICATION problem}
%
There are a large number of learning algorithms to choose from, existing
on a spectrum that balances expressiveness with ease of training (and of
interpreting the learned model).
%
In this section we present the learning algorithms we will investigate,
beginning with the simplest.
%
For each algorithm, we describe \ES{TODO...}

@@

Linear models like logistic regression are popular for their ease of
training and of interpreting the resulting model; however, they can be
limited in their applicability.
%
As the name implies, linear models may perform poorly when asked to
learn an inherently \emph{nonlinear} function.
%
This limitation can be mitigated to some extent by adding more features
to the model (\eg combinations of the existing features \ES{CITE?}), but
the fundamental issue remains.

>>>>

\section{Learning to Blame}
\label{sec:learning}
In this section, we describe our approach to localizing type errors, in the
context
of \lang (\autoref{fig:syntax}), a simple lambda calculus with integers,
booleans, pairs, and lists.
%
\input{syntax}
%
First, we define the inputs to the model, a set of \emph{features}
(\autoref{sec:features}) that we will use to describe programs.
%
Second, we define the expected outputs of the model, a set of
\emph{labels} (\autoref{sec:labels}) that we will use to assign blame.
%
Third, we describe the actual \emph{models} (\autoref{sec:models}) that we
will evaluate.

@@

% \begin{figure}[ht]
% \begin{minipage}{0.6\linewidth}
% \begin{code}
%   let rec sumList xs =
%     match xs with
%     | []     -> []
%     | hd::tl -> hd + sumList tl
% \end{code}
% \end{minipage}
% \begin{minipage}{0.3\linewidth}
% \begin{code}[numbers=left]
% []
% hd + sumList tl
% sumList tl
% tl
% \end{code}
% \end{minipage}
% \caption{An ill-typed program (left) and a selection of its
%   sub-expressions (right).}
% \label{fig:sumList}
% \end{figure}
\begin{table}[ht]
\begin{tabular}{lrrrrrr}
\toprule
Expression
  & \IsNil & \IsCaseListP & \ExprSize
  & \HasTypeIntCOne & \HasTypeList & \InSlice \\
\midrule
|[]|
  & 1 & 1 & 1 & 0 & 1 & 1 \\
|hd + sumList tl|
  & 0 & 1 & 5 & 1 & 0 & 1 \\
|sumList tl|
  & 0 & 0 & 3 & 0 & 1 & 1 \\
|tl|
  & 0 & 0 & 1 & 0 & 1 & 0 \\
\bottomrule
\end{tabular}
\caption{A selection of the features we would extract from the
  \lstinline!sumList! program in \autoref{fig:sumList}. A feature is
  considered \emph{enabled} if it has a non-zero value, and
  \emph{disabled} otherwise. A ``-P'' suffix indicates that the feature
  describes the parent of the current expression, a ``-C$n$'' suffix
  indicates that the feature describes the $n$-th (left-to-right) child
  of the current expression.  Note that, since we rely on a partial
  typing derivation, we are subject to the well-known traversal bias and
  label the expression \lstinline!sumList tl! as having type
  $\tlist{\cdot}$. The model will have to learn to correct for this
  bias.}
\label{tab:sumList}
\end{table}

@@

We distinguish between matching on a list vs.\ on a pair, as this
affects the typing derivation.
%
We also assume that all pattern matches are well-formed --- \ie all
patterns must match on the same type.
%
Ill-formed match expressions would lead to a type error; however, they
are already effectively localized to the match expression itself.
%
We note that this is not a \emph{fundamental} limitation, and one could
easily add features that specify whether a match \emph{contains} a
particular pattern, and thus have a match expression that enables multiple
features.

\subsubsection{Contextual syntactic features}
These are similar to local syntactic features, but lifted to describe the
parent and children of the current expression.
%
For example, the \IsCaseListP feature in \autoref{tab:sumList} describes
whether current expression's \emph{parent} matches on a list.
%
If a particular $e$ does not have children (\eg a variable $x$) or a
parent (\ie the root expression), we leave the corresponding features
disabled.
%
This gives us a notion of the \emph{context} in which an expression
occurs, similar to the \emph{n-grams} commonly used in linguistic
models~\cite{FIXME,FIXME}.
% FIXME: cite "on the naturalness of software" and "on the uniqueness of
% software" at the very least for the above

@@

\subsubsection{Expression size}
We also propose a feature representing the \emph{size} of each expression,
\ie how many sub-expressions does it contain?
%
For example, the \ExprSize feature in \autoref{tab:sumList} is set to three
for the expression |sumList tl| as it contains three expressions:
the two variables and the application itself.
%
This allows the model to learn that, \eg, expressions closer to the
leaves are more likely to be blamed than expressions closer to the root.

@@

Note that our use of typing features in an ill-typed program
subjects us to \emph{traversal
bias}~\cite{McAdam1998-ub,Yang2000-yr}. For example, the 
|sumList tl| expression might alternatively be assigned
the type $\tint$.
%
We rely on the model to counteract this bias (see FIXME reference to later
in the paper).

\subsubsection{Type error slice}
Finally, we desire a model that distinguishes between
changes that could fix the error, and changes that
\emph{cannot possibly} fix the error.
%
Thus, we compute a minimal type error \emph{slice} for the program
(\ie the set of expressions that contribute to the error), and add a
feature that is enabled for expressions that are part of the slice.
%
The \InSlice feature in \autoref{tab:sumList} indicates whether an
expression is part of such a minimal slice, and is enabled for all of
the sampled expressions except for |tl|, which does not affect
the type error.
%
If the program contains multiple type errors, we compute
a minimal slice for each error.

@@

\subsection{Learning Algorithms}
\label{sec:models}
\ES{TODO: mention that this is a SUPERVISED, CLASSIFICATION problem}
%
There are many learning algorithms to choose from, existing
on a spectrum that balances expressiveness with ease of training (and of
interpreting the learned model).
%
In this section we present the learning algorithms we will investigate,
beginning with the simplest.
%
For each algorithm, we describe \ES{TODO...}

@@

Linear models, like logistic regression, are popular for their ease of
% FIXME: Are you sure you want to call LOGISTIC regression a LINEAR model? 
training and of interpreting the resulting model; however, they can be
limited in their applicability.
%
As the name implies, linear models may perform poorly when asked to
learn an inherently \emph{nonlinear} function.
%
This limitation can be mitigated to some extent by adding more features
to the model (\eg combinations of the existing features \ES{CITE?}), but
the fundamental issue remains.
